---
title: "AlphaGenome GRCh38 Enhancer-versus-Silencer Comparison"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
params:
  evn_results_dir: null
  svn_results_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5.5, dpi = 120)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(gt)
  library(scales)
})

learning_fractions <- c(10, 20, 40, 60, 80, 100)
outer_names <- paste0("outer", 1:4)
task_colors <- c("Enhancer vs negative" = "#0072B2",
                 "Silencer vs negative" = "#D55E00")
theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom"))
```

```{r inputs, include=FALSE}
read_metrics <- function(path, comparison) {
  metrics_path <- file.path(normalizePath(path, mustWork = TRUE), "metrics.tsv")
  if (!file.exists(metrics_path)) stop("Missing metrics table: ", metrics_path)
  x <- read.delim(metrics_path, sep = "\t", stringsAsFactors = FALSE,
                  check.names = FALSE)
  required <- c("model", "task", "dataset", "dataset_type", "training_fraction",
                "split", "n_training_total", "n_evaluation_total", "best_epoch",
                "elapsed_seconds", "auroc", "auprc", "accuracy", "precision",
                "recall", "f1")
  missing <- setdiff(required, names(x))
  if (length(missing)) stop("Missing columns in ", metrics_path, ": ", paste(missing, collapse = ", "))
  if (nrow(x) != 20) stop("Expected 20 rows in ", metrics_path)
  if (!all(x$model == "alphagenome") || !all(x$task == "GRCh38")) {
    stop("Model/task mismatch in ", metrics_path)
  }
  if (!identical(sort(unique(x$dataset)), sort(c(paste0("learning_curve_", learning_fractions), outer_names)))) {
    stop("Dataset inventory mismatch in ", metrics_path)
  }
  x |> mutate(comparison = comparison)
}

if (is.null(params$evn_results_dir) || is.null(params$svn_results_dir)) {
  stop("Both EVN and SVN result directories are required")
}

metrics <- bind_rows(
  read_metrics(params$evn_results_dir, "Enhancer vs negative"),
  read_metrics(params$svn_results_dir, "Silencer vs negative")
) |>
  mutate(comparison = factor(comparison, levels = names(task_colors)),
         split = factor(split, levels = c("validation", "test"),
                        labels = c("Validation", "Test")),
         ranking_error = 1 - auroc)

learning <- metrics |>
  filter(dataset_type == "learning_curve") |>
  arrange(split, comparison, n_training_total)
outer <- metrics |>
  filter(dataset_type == "outer") |>
  mutate(dataset = factor(dataset, levels = outer_names))
```

# Scope and interpretation

This report compares learning behavior and chromosome-composition stability for two
frozen-encoder AlphaGenome classifiers: enhancer versus matched synthetic negative
sequences (EVN), and silencer versus matched synthetic negative sequences (SVN). Both
use the same encoder, classification-head architecture, optimization configuration,
checkpoint-selection rule, and six nested percentage checkpoints.

The comparison is descriptive rather than paired. EVN and SVN contain different
positive loci and different validation/test cohorts, and their 100% training pools
are not exactly the same size. Vertical differences may therefore reflect both
biological task difficulty and cohort composition. Absolute training-observation
counts are used on the x-axis.

```{r training-size-table}
learning |>
  filter(split == "Validation") |>
  distinct(comparison, training_fraction, n_training_total) |>
  mutate(`Training fraction` = paste0(training_fraction, "%")) |>
  select(Task = comparison, `Training fraction`,
         `Training observations` = n_training_total) |>
  gt(groupname_col = "Task") |>
  tab_header(title = "Task-specific learning-curve sizes") |>
  fmt_integer(columns = `Training observations`)
```

# AUROC learning curves

```{r auroc-curves}
ggplot(learning,
       aes(n_training_total, auroc, color = comparison,
           shape = comparison, group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~split, scales = "free_y") +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.14, 0.18))) +
  labs(title = "AlphaGenome AUROC learning curves",
       x = "Training observations", y = "AUROC", color = NULL, shape = NULL)
```

# Average-precision learning curves

```{r ap-curves}
ggplot(learning,
       aes(n_training_total, auprc, color = comparison,
           shape = comparison, group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~split, scales = "free_y") +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.14, 0.18))) +
  labs(title = "AlphaGenome Average-Precision learning curves",
       x = "Training observations", y = "Average Precision",
       color = NULL, shape = NULL)
```

# Pairwise AUROC ranking error

`1 - AUROC` is the pairwise ranking error implied by AUROC, not thresholded
classification error.

```{r ranking-error-curves}
ggplot(learning,
       aes(n_training_total, ranking_error, color = comparison,
           shape = comparison, group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~split, scales = "free_y") +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.14, 0.18))) +
  labs(title = "AlphaGenome pairwise AUROC ranking-error curves",
       x = "Training observations", y = "1 - AUROC",
       color = NULL, shape = NULL)
```

# Chromosome-composition stability

Outer-set numbering is task-specific because the chromosome groupings were balanced
separately for EVN and SVN. The plots compare each task's stability pattern, not the
same outer training composition across tasks.

```{r outer-stability}
outer_long <- outer |>
  filter(split == "Test") |>
  select(comparison, dataset, auroc, auprc) |>
  pivot_longer(c(auroc, auprc), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric, auroc = "AUROC", auprc = "Average Precision"))

ggplot(outer_long,
       aes(dataset, value, color = comparison, shape = comparison,
           group = comparison)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  facet_wrap(~metric, scales = "free_y") +
  scale_color_manual(values = task_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.16, 0.18))) +
  labs(title = "Held-out outer-test stability",
       x = "Task-specific outer set", y = NULL, color = NULL, shape = NULL)
```

```{r outer-summary}
outer |>
  filter(split == "Test") |>
  group_by(comparison) |>
  summarise(`Mean AUROC` = mean(auroc), `SD AUROC` = sd(auroc),
            `Minimum AUROC` = min(auroc), `Maximum AUROC` = max(auroc),
            `Mean AP` = mean(auprc), `SD AP` = sd(auprc),
            `Minimum AP` = min(auprc), `Maximum AP` = max(auprc), .groups = "drop") |>
  rename(Task = comparison) |>
  gt() |>
  tab_header(title = "Held-out outer-test summary") |>
  fmt_number(columns = where(is.numeric), decimals = 4)
```

# Runtime comparison

Elapsed time covers model loading, training, validation, checkpoint restoration, and
test evaluation. Both tasks ran on NVIDIA A100-SXM4-40GB GPUs with the same batch size
and maximum epoch count.

```{r runtime-curves}
runtime <- learning |>
  filter(split == "Test") |>
  mutate(elapsed_minutes = elapsed_seconds / 60)

ggplot(runtime,
       aes(n_training_total, elapsed_minutes, color = comparison,
           shape = comparison, group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma) +
  labs(title = "End-to-end AlphaGenome runtime",
       x = "Training observations", y = "Elapsed minutes",
       color = NULL, shape = NULL)
```

# Full-training comparison

```{r full-training-table}
learning |>
  filter(training_fraction == 100) |>
  select(Task = comparison, Split = split,
         `Training observations` = n_training_total,
         `Evaluation observations` = n_evaluation_total,
         `Best epoch` = best_epoch, Accuracy = accuracy,
         AUROC = auroc, `Average Precision` = auprc,
         Precision = precision, Recall = recall, F1 = f1) |>
  gt(groupname_col = "Task") |>
  tab_header(title = "100% AlphaGenome model performance") |>
  fmt_integer(columns = c(`Training observations`, `Evaluation observations`, `Best epoch`)) |>
  fmt_number(columns = c(Accuracy, AUROC, `Average Precision`, Precision, Recall, F1),
             decimals = 4)
```

# Reproducibility notes

- Both inputs are the validated `metrics.tsv` tables produced by
  `src/prepare_alphagenome_results.py`.
- Each percentage was trained independently from a newly initialized head.
- Validation AUPRC selected every checkpoint; test data were evaluated afterward.
- Aggregate artifacts do not contain per-sequence logits, so full ROC/PR traces and
  top-fraction enrichment cannot be reconstructed here.

Rendered at `r format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`.

```{r session-info}
sessionInfo()
```
