---
title: "AlphaGenome GRCh38 Evaluation and Learning-Curve Report"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
params:
  results_dir: null
  analysis_name: null
  positive_class: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 8.5, fig.height = 5.5, dpi = 120)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(gt)
  library(scales)
})

learning_fractions <- c(10, 20, 40, 60, 80, 100)
outer_names <- paste0("outer", 1:4)
split_colors <- c(Validation = "#D55E00", Test = "#0072B2")
outer_colors <- c(outer1 = "#0072B2", outer2 = "#D55E00",
                  outer3 = "#009E73", outer4 = "#CC79A7")
theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom"))
```

```{r inputs, include=FALSE}
results_dir <- params$results_dir
analysis_name <- params$analysis_name
positive_class <- params$positive_class
if (is.null(results_dir) || !nzchar(results_dir)) stop("params$results_dir is required")
if (is.null(analysis_name) || !nzchar(analysis_name)) stop("params$analysis_name is required")
if (is.null(positive_class) || !nzchar(positive_class)) stop("params$positive_class is required")
results_dir <- normalizePath(results_dir, mustWork = TRUE)
metrics_path <- file.path(results_dir, "metrics.tsv")
history_path <- file.path(results_dir, "history.tsv")
if (!file.exists(metrics_path) || !file.exists(history_path)) {
  stop("Missing AlphaGenome aggregate input under ", results_dir)
}

metrics <- read.delim(metrics_path, sep = "\t", stringsAsFactors = FALSE,
                      check.names = FALSE)
history <- read.delim(history_path, sep = "\t", stringsAsFactors = FALSE,
                      check.names = FALSE)

required_metrics <- c(
  "model", "task", "dataset", "dataset_type", "training_fraction", "split",
  "n_training_positive", "n_training_negative", "n_training_total",
  "n_evaluation_positive", "n_evaluation_negative", "n_evaluation_total",
  "best_epoch", "epochs_completed", "selection_metric", "elapsed_seconds",
  "loss", "accuracy", "auroc", "auprc", "precision", "recall", "f1"
)
missing_metrics <- setdiff(required_metrics, names(metrics))
if (length(missing_metrics)) stop("Missing metrics columns: ", paste(missing_metrics, collapse = ", "))
if (nrow(metrics) != 20) stop("Expected 20 metric rows; found ", nrow(metrics))
if (!identical(sort(unique(metrics$dataset)), sort(c(paste0("learning_curve_", learning_fractions), outer_names)))) {
  stop("Unexpected AlphaGenome dataset inventory")
}
if (!all(metrics$model == "alphagenome") || !all(metrics$task == "GRCh38")) {
  stop("Model/task metadata mismatch")
}
if (!all(metrics$selection_metric == "validation_auprc")) {
  stop("Unexpected checkpoint selection metric")
}
if (!all(metrics$split %in% c("validation", "test"))) stop("Invalid split")
if (anyNA(metrics$auroc) || anyNA(metrics$auprc)) stop("Missing ranking metric")

metrics <- metrics |>
  mutate(split = factor(split, levels = c("validation", "test"),
                        labels = c("Validation", "Test")),
         ranking_error = 1 - auroc)
learning_metrics <- metrics |>
  filter(dataset_type == "learning_curve") |>
  arrange(split, n_training_total)
outer_metrics <- metrics |>
  filter(dataset_type == "outer") |>
  mutate(dataset = factor(dataset, levels = outer_names)) |>
  arrange(split, dataset)
test_learning <- learning_metrics |> filter(split == "Test") |> arrange(n_training_total)

training_table <- learning_metrics |>
  filter(split == "Validation") |>
  select(training_fraction, n_training_positive, n_training_negative, n_training_total)

config_fields <- c("sequence_length", "hidden_size", "dropout", "batch_size", "epochs",
                   "learning_rate", "weight_decay", "patience", "seed", "threshold",
                   "max_shift", "rc_prob", "shift_prob", "device_name", "torch_version",
                   "numpy_version", "weights")
for (field in config_fields) {
  if (length(unique(metrics[[field]])) != 1) stop("Configuration mismatch: ", field)
}
config <- metrics[1, config_fields, drop = FALSE]
```

# Dataset and experimental design

The GRCh38 `r analysis_name` experiment uses a frozen pretrained AlphaGenome
encoder and an independently initialized binary classification head for every run.
`r positive_class` sequences are label 1 and matched synthetic negatives are label 0.

Two designs are reported separately. The main learning curve uses six strictly nested
training subsets with one fixed chr1 validation set and one fixed held-out chr2 test
set. The four chromosome-composition stability experiments use outer-specific training
and validation chromosomes while reserving fold 5 as their common held-out test set.
Validation AUPRC selects the checkpoint and controls early stopping. The test split is
evaluated only after the selected checkpoint has been restored.

```{r training-size-table}
training_table |>
  mutate(`Training fraction` = paste0(training_fraction, "%")) |>
  select(`Training fraction`,
         `Training positives` = n_training_positive,
         `Training negatives` = n_training_negative,
         `Total training observations` = n_training_total) |>
  gt() |>
  tab_header(title = "Strictly nested learning-curve training sets") |>
  fmt_integer(columns = c(`Training positives`, `Training negatives`,
                          `Total training observations`))
```

```{r evaluation-size-table}
metrics |>
  distinct(dataset_type, dataset, split, n_evaluation_positive,
           n_evaluation_negative, n_evaluation_total) |>
  mutate(Design = if_else(dataset_type == "learning_curve", "Fixed learning curve", "Outer stability")) |>
  select(Design, Dataset = dataset, Split = split,
         Positives = n_evaluation_positive, Negatives = n_evaluation_negative,
         `Total observations` = n_evaluation_total) |>
  gt(groupname_col = "Design") |>
  tab_header(title = "Validation and test-set sizes") |>
  fmt_integer(columns = c(Positives, Negatives, `Total observations`))
```

# Model and training configuration

The pretrained AlphaGenome trunk is loaded without its original output heads and kept
fully frozen. Encoder tensors are normalized to `(batch, positions, 1536)`; in this
installation a 256-bp input produces two encoder positions. The classifier applies
layer normalization, flattens the encoder representation, and uses a
`3072 → 1024 → 1` MLP with ReLU and dropout. Its scalar output is an unbounded logit,
optimized with binary cross-entropy with logits.

```{r model-metadata}
tribble(
  ~Setting, ~Value,
  "Pretrained weights", as.character(config$weights),
  "Encoder", "Pretrained AlphaGenome trunk; frozen",
  "Input sequence length", as.character(config$sequence_length),
  "Encoder representation", "2 positions × 1536 channels",
  "Classification head", paste0("LayerNorm; flatten; Linear(3072, ", config$hidden_size,
                                "); ReLU; dropout; Linear(", config$hidden_size, ", 1)"),
  "Loss", "Binary cross-entropy with logits",
  "Positive class", paste0(positive_class, " (label 1)"),
  "Dropout", as.character(config$dropout),
  "Batch size", as.character(config$batch_size),
  "Maximum epochs", as.character(config$epochs),
  "Learning rate", as.character(config$learning_rate),
  "Weight decay", as.character(config$weight_decay),
  "Early-stopping patience", as.character(config$patience),
  "Checkpoint selection", "Maximum validation AUPRC",
  "Random seed", as.character(config$seed),
  "Augmentation", paste0("reverse-complement p=", config$rc_prob,
                         "; shift p=", config$shift_prob,
                         "; maximum shift=", config$max_shift, " bp"),
  "Mixed precision", "CUDA bfloat16 autocast",
  "GPU", as.character(config$device_name),
  "PyTorch", as.character(config$torch_version),
  "NumPy", as.character(config$numpy_version)
) |>
  gt() |>
  tab_header(title = "Frozen AlphaGenome training configuration")
```

# Chromosome-composition stability

The outer experiments assess sensitivity to chromosome composition. Their validation
sets differ by outer design, whereas every outer model is evaluated on the same held-out
fold 5 test data. These are descriptive stability comparisons, not pooled cross-validation
estimates.

```{r outer-ranking-plots}
outer_long <- outer_metrics |>
  select(dataset, split, auroc, auprc) |>
  pivot_longer(c(auroc, auprc), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric, auroc = "AUROC", auprc = "Average Precision"))

ggplot(outer_long,
       aes(dataset, value, color = split, shape = split, group = split)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  facet_wrap(~metric, scales = "free_y") +
  scale_color_manual(values = split_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.18))) +
  labs(title = "Validation and held-out test ranking performance across outer sets",
       x = "Outer set", y = NULL, color = "Split", shape = "Split")
```

```{r outer-table}
outer_metrics |>
  select(`Outer set` = dataset, Split = split,
         `Evaluation observations` = n_evaluation_total,
         `Best epoch` = best_epoch, AUROC = auroc,
         `Average Precision` = auprc, `Ranking error` = ranking_error) |>
  gt(groupname_col = "Outer set") |>
  fmt_integer(columns = c(`Evaluation observations`, `Best epoch`)) |>
  fmt_number(columns = c(AUROC, `Average Precision`, `Ranking error`), decimals = 4)
```

```{r outer-summary, include=FALSE}
outer_test <- outer_metrics |> filter(split == "Test")
outer_auroc_mean <- mean(outer_test$auroc)
outer_ap_mean <- mean(outer_test$auprc)
```

Held-out fold 5 AUROC averages `r sprintf("%.4f", outer_auroc_mean)` and spans
`r sprintf("%.4f–%.4f", min(outer_test$auroc), max(outer_test$auroc))`. Average
precision averages `r sprintf("%.4f", outer_ap_mean)` and spans
`r sprintf("%.4f–%.4f", min(outer_test$auprc), max(outer_test$auprc))`.

# Learning-curve analysis

Absolute training observations define the x-axis. Validation and held-out test results
are overlaid because every learning-curve model uses the same evaluation loci.

## AUROC

```{r learning-auroc}
ggplot(learning_metrics,
       aes(n_training_total, auroc, color = split, shape = split, group = split)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  geom_text(data = filter(learning_metrics, split == "Test"),
            aes(label = sprintf("%d%%\n%.3f", training_fraction, auroc)),
            vjust = -0.8, size = 3, show.legend = FALSE) +
  geom_text(data = filter(learning_metrics, split == "Validation"),
            aes(label = sprintf("%d%%\n%.3f", training_fraction, auroc)),
            vjust = 1.8, size = 3, show.legend = FALSE) +
  scale_color_manual(values = split_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.06))) +
  scale_y_continuous(expand = expansion(mult = c(0.22, 0.25))) +
  labs(title = "Validation and test AUROC by absolute training-set size",
       x = "Training observations", y = "AUROC", color = "Split", shape = "Split")
```

## Average precision

```{r learning-ap}
ggplot(learning_metrics,
       aes(n_training_total, auprc, color = split, shape = split, group = split)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  geom_text(data = filter(learning_metrics, split == "Test"),
            aes(label = sprintf("%d%%\n%.3f", training_fraction, auprc)),
            vjust = -0.8, size = 3, show.legend = FALSE) +
  geom_text(data = filter(learning_metrics, split == "Validation"),
            aes(label = sprintf("%d%%\n%.3f", training_fraction, auprc)),
            vjust = 1.8, size = 3, show.legend = FALSE) +
  scale_color_manual(values = split_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.06))) +
  scale_y_continuous(expand = expansion(mult = c(0.22, 0.25))) +
  labs(title = "Validation and test Average Precision by absolute training-set size",
       x = "Training observations", y = "Average Precision",
       color = "Split", shape = "Split")
```

## Pairwise AUROC ranking error

`1 - AUROC` is the pairwise AUROC ranking error: approximately the probability that
a positive-negative pair is ordered incorrectly, with ties handled according to the
AUROC convention. It is not thresholded classification error.

```{r ranking-error}
ggplot(learning_metrics,
       aes(n_training_total, ranking_error, color = split,
           shape = split, group = split)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  scale_color_manual(values = split_colors) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(expand = expansion(mult = c(0.12, 0.14))) +
  labs(title = "Validation and test pairwise AUROC ranking error",
       x = "Training observations", y = "1 - AUROC",
       color = "Split", shape = "Split")
```

## Marginal held-out test gains

```{r marginal-gains}
marginal_gains <- test_learning |>
  transmute(training_fraction, n_training_total, auroc, auprc, ranking_error) |>
  mutate(from_fraction = lag(training_fraction), from_n = lag(n_training_total),
         delta_n = n_training_total - from_n,
         delta_auroc = auroc - lag(auroc),
         delta_ap = auprc - lag(auprc),
         ranking_error_reduction = lag(ranking_error) - ranking_error,
         delta_auroc_per_10000 = delta_auroc / delta_n * 10000) |>
  filter(!is.na(from_fraction)) |>
  transmute(From = paste0(from_fraction, "%"), To = paste0(training_fraction, "%"),
            `Additional training observations` = delta_n,
            `Change in AUROC` = delta_auroc, `Change in AP` = delta_ap,
            `Reduction in ranking error` = ranking_error_reduction,
            `AUROC gain per 10,000 observations` = delta_auroc_per_10000)

marginal_gains |>
  gt() |>
  fmt_integer(columns = `Additional training observations`) |>
  fmt_number(columns = c(`Change in AUROC`, `Change in AP`,
                         `Reduction in ranking error`,
                         `AUROC gain per 10,000 observations`), decimals = 4)
```

# Training dynamics and early stopping

Each percentage starts from a newly initialized classification head. The curves below
show the per-epoch validation AUPRC used for checkpoint selection; points identify the
restored best epoch. Runs can contain more epochs than the selected checkpoint because
training continues until the patience criterion is reached or the ten-epoch maximum is
completed.

```{r history-auprc}
learning_history <- history |>
  filter(dataset_type == "learning_curve") |>
  mutate(training_fraction = factor(training_fraction, levels = learning_fractions,
                                    labels = paste0(learning_fractions, "%")))
best_history <- learning_history |> filter(epoch == best_epoch)

ggplot(learning_history,
       aes(epoch, validation_auprc, color = training_fraction,
           group = training_fraction)) +
  geom_line(linewidth = 0.85) +
  geom_point(data = best_history, size = 2.8) +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = seq_len(max(history$epoch))) +
  labs(title = "Validation AUPRC during learning-curve training",
       subtitle = "Points mark validation-selected checkpoints",
       x = "Epoch", y = "Validation AUPRC", color = "Training fraction")
```

```{r history-loss}
loss_history <- learning_history |>
  select(dataset, training_fraction, epoch, train_loss, validation_loss) |>
  pivot_longer(c(train_loss, validation_loss), names_to = "split", values_to = "loss") |>
  mutate(split = recode(split, train_loss = "Training", validation_loss = "Validation"))

ggplot(loss_history, aes(epoch, loss, color = split, group = split)) +
  geom_line(linewidth = 0.75) +
  facet_wrap(~training_fraction, scales = "free_y") +
  scale_color_manual(values = c(Training = "#009E73", Validation = "#D55E00")) +
  scale_x_continuous(breaks = c(1, 5, 10)) +
  labs(title = "Training and validation loss by epoch",
       x = "Epoch", y = "Binary cross-entropy", color = NULL)
```

# Runtime scaling

Elapsed time includes model loading, training, repeated validation, restoration of the
best checkpoint, and final test evaluation. It is therefore an end-to-end run time,
not training-only GPU time.

```{r runtime}
runtime_data <- learning_metrics |>
  filter(split == "Test") |>
  mutate(elapsed_minutes = elapsed_seconds / 60)

ggplot(runtime_data, aes(n_training_total, elapsed_minutes)) +
  geom_line(color = "#0072B2", linewidth = 0.9) +
  geom_point(color = "#0072B2", size = 2.8) +
  geom_text(aes(label = sprintf("%d%%\n%.1f min", training_fraction, elapsed_minutes)),
            vjust = -0.7, size = 3) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.06))) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.22))) +
  labs(title = "End-to-end runtime by training-set size",
       x = "Training observations", y = "Elapsed minutes")
```

# Final-model metrics

AUROC and average precision are threshold-independent ranking metrics. Accuracy,
precision, recall, and F1 below use `sigmoid(logit) ≥ 0.5`, the fixed threshold encoded
in the training configuration; it was not selected on the test set.

```{r final-metrics}
learning_metrics |>
  filter(training_fraction == 100) |>
  select(Split = split, `Evaluation observations` = n_evaluation_total,
         `Best epoch` = best_epoch, Loss = loss, Accuracy = accuracy,
         AUROC = auroc, `Average Precision` = auprc,
         Precision = precision, Recall = recall, F1 = f1) |>
  gt() |>
  tab_header(title = "100% AlphaGenome model performance") |>
  fmt_integer(columns = c(`Evaluation observations`, `Best epoch`)) |>
  fmt_number(columns = c(Loss, Accuracy, AUROC, `Average Precision`,
                         Precision, Recall, F1), decimals = 4)
```

# Interpretation

```{r interpretation-values, include=FALSE}
smallest <- slice_head(test_learning, n = 1)
largest <- slice_tail(test_learning, n = 1)
best_test <- test_learning |> slice_max(auroc, n = 1, with_ties = FALSE)
```

From `r comma(smallest$n_training_total)` to `r comma(largest$n_training_total)`
training observations, held-out test AUROC changes from
`r sprintf("%.4f", smallest$auroc)` to `r sprintf("%.4f", largest$auroc)`, and
average precision changes from `r sprintf("%.4f", smallest$auprc)` to
`r sprintf("%.4f", largest$auprc)`. The highest observed test AUROC is
`r sprintf("%.4f", best_test$auroc)` at `r paste0(best_test$training_fraction, "%")`.
The small difference between the 80% and 100% points should not be treated as evidence
that additional data is intrinsically harmful: each model is a stochastic independent
fit, and there is one run per training size rather than repeated seeds.

The narrow fold 5 performance range across outer models supports chromosome-composition
stability for this design. It does not estimate uncertainty from model initialization,
which would require repeated seeds.

# Scope limitation

The saved AlphaGenome artifacts contain aggregate metrics and epoch histories, not
per-sequence logits. Consequently, this report cannot reconstruct full ROC or
precision-recall traces, confusion matrices, alternative thresholds, or top-1/5/10%
enrichment. Those analyses would require either retaining sequence-level scores or
computing the desired summaries inside the evaluator. This does not affect the main
AUROC/AUPRC learning-curve analysis presented here.

# Reproducibility

```{r artifact-summary}
metrics |>
  distinct(dataset, dataset_type, n_training_total, best_epoch, epochs_completed,
           elapsed_seconds) |>
  mutate(`Elapsed minutes` = elapsed_seconds / 60,
         Design = recode(dataset_type,
                         learning_curve = "Fixed learning curve",
                         outer = "Outer stability")) |>
  select(Design, Dataset = dataset, `Training observations` = n_training_total,
         `Best epoch` = best_epoch, `Epochs completed` = epochs_completed,
         `Elapsed minutes`) |>
  gt(groupname_col = "Design") |>
  fmt_integer(columns = c(`Training observations`, `Best epoch`, `Epochs completed`)) |>
  fmt_number(columns = `Elapsed minutes`, decimals = 1)
```

The report reads `metrics.tsv` and `history.tsv`, generated by
`src/prepare_alphagenome_results.py`. That script validates the complete five-artifact
set for every run, verifies one frozen configuration, checks validation-AUPRC checkpoint
selection and restored epoch consistency, and counts the FASTA records referenced by
each run configuration.

Rendered at `r format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`.

```{r session-info}
sessionInfo()
```
