---
title: "GRCh38 Native-versus-Cross-Task Evaluation"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
params:
  metrics: null
  model: null
  feature_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.width=9, fig.height=5.8, dpi=120)
suppressPackageStartupMessages({library(dplyr); library(tidyr); library(ggplot2); library(scales); library(gt)})
if (is.null(params$metrics) || is.null(params$model)) stop("metrics and model parameters are required")
all_metrics <- read.delim(params$metrics, check.names=FALSE)
x <- all_metrics |> filter(model == params$model) |>
  mutate(`Trained task`=recode(model_task, evn="EVN", svn="SVN"),
         `Evaluation task`=recode(evaluation_task, evn="EVN", svn="SVN"),
         Split=factor(recode(split, validation="Validation", test="Test"),
                      levels=c("Validation","Test")),
         Mode=factor(recode(evaluation_mode, native="Native", cross="Cross-task"),
                     levels=c("Native","Cross-task")))
if (nrow(x) != 48) stop("Expected 48 metric rows for one model family")
model_name <- if (params$model == "alphagenome") "AlphaGenome" else "LS-GKM"
auroc_limits <- if (params$model == "alphagenome") c(0.785,0.835) else c(0.67,0.725)
auprc_limits <- if (params$model == "alphagenome") c(0.77,0.825) else c(0.635,0.695)
colors <- c(EVN="#0072B2", SVN="#D55E00")
theme_set(theme_minimal(base_size=12) + theme(panel.grid.minor=element_blank(), legend.position="bottom"))
```

# Scope and interpretation

This report evaluates each `r model_name` learning-curve model on both its native
binary task and the other regulatory task. EVN-trained models are transferred to
SVN validation/test data, and SVN-trained models are transferred to EVN data.

Cross-task AUROC/AP measures positive-versus-negative transfer. It does not measure
direct enhancer-versus-silencer discrimination. The tasks have different positive
loci and independently generated negative backgrounds, so differences remain
descriptive rather than a paired causal comparison.

```{r evaluation-sizes}
x |> distinct(`Evaluation task`, Split, n_positive, n_negative, n_evaluation_total) |>
  select(`Evaluation task`, Split, Positives=n_positive, Negatives=n_negative,
         Total=n_evaluation_total) |> gt() |> fmt_integer(columns=c(Positives,Negatives,Total))
```

# Learning curves

## AUROC

```{r auroc}
ggplot(x, aes(n_training_total, auroc, color=`Evaluation task`, linetype=Mode,
              shape=Mode, group=interaction(`Evaluation task`,Mode))) +
  geom_line(linewidth=.9) + geom_point(size=2.5) +
  facet_grid(Split ~ `Trained task`) + scale_color_manual(values=colors) +
  scale_linetype_manual(values=c(Native="solid", `Cross-task`="dashed")) +
  scale_shape_manual(values=c(Native=16, `Cross-task`=17)) +
  scale_x_continuous(labels=comma) +
  scale_y_continuous(limits=auroc_limits, breaks=pretty_breaks(6)) +
  labs(title=paste(model_name,"native and cross-task AUROC"),
       x="Training observations", y="AUROC", color="Evaluation data", linetype="Mode") +
  guides(shape="none",
         linetype=guide_legend(override.aes=list(shape=NA, color="black", linewidth=1)))
```

## Average precision

```{r auprc}
ggplot(x, aes(n_training_total, auprc, color=`Evaluation task`, linetype=Mode,
              shape=Mode, group=interaction(`Evaluation task`,Mode))) +
  geom_line(linewidth=.9) + geom_point(size=2.5) +
  facet_grid(Split ~ `Trained task`) + scale_color_manual(values=colors) +
  scale_linetype_manual(values=c(Native="solid", `Cross-task`="dashed")) +
  scale_shape_manual(values=c(Native=16, `Cross-task`=17)) +
  scale_x_continuous(labels=comma) +
  scale_y_continuous(limits=auprc_limits, breaks=pretty_breaks(6)) +
  labs(title=paste(model_name,"native and cross-task average precision"),
       x="Training observations", y="Average Precision", color="Evaluation data", linetype="Mode") +
  guides(shape="none",
         linetype=guide_legend(override.aes=list(shape=NA, color="black", linewidth=1)))
```

```{r feature-inputs, include=FALSE}
has_features <- params$model == "alphagenome" && !is.null(params$feature_dir)
if (has_features) {
  feature_importance <- read.delim(file.path(params$feature_dir,"feature_importance.tsv"))
  feature_similarity <- read.delim(file.path(params$feature_dir,"feature_similarity.tsv"))
}
```

```{r feature-section, results='asis'}
if (has_features) cat("# AlphaGenome head-feature attribution\n\n")
```

```{r feature-explanation, results='asis'}
if (has_features) cat("The EVN and SVN heads receive the same 3,072 aligned encoder dimensions. Importance is the summed absolute input-to-hidden-to-output path strength, adjusted by the LayerNorm scale. This measures global connection capacity. It does not include sequence-dependent ReLU gating and does not assign biological names to latent encoder channels.\n\n")
```

```{r feature-similarity, eval=has_features}
similarity_long <- feature_similarity |>
  select(training_fraction, cosine_similarity, spearman_correlation) |>
  pivot_longer(-training_fraction, names_to="metric", values_to="value") |>
  mutate(metric=recode(metric, cosine_similarity="Cosine similarity",
                       spearman_correlation="Spearman correlation"))
ggplot(similarity_long, aes(training_fraction, value, color=metric)) +
  geom_line(linewidth=.9) + geom_point(size=2.5) +
  scale_x_continuous(breaks=c(10,20,40,60,80,100), labels=paste0(c(10,20,40,60,80,100),"%")) +
  scale_y_continuous(limits=c(0,1)) +
  labs(title="Similarity of EVN and SVN head importance profiles",
       x="Training fraction", y="Similarity", color=NULL)
```

```{r top-overlap, eval=has_features}
overlap <- feature_similarity |>
  select(training_fraction, ends_with("overlap_fraction")) |>
  pivot_longer(-training_fraction, names_to="cutoff", values_to="overlap") |>
  mutate(cutoff=recode(cutoff, top_1_overlap_fraction="Top 1%",
                       top_5_overlap_fraction="Top 5%", top_10_overlap_fraction="Top 10%"))
ggplot(overlap, aes(training_fraction, overlap, color=cutoff)) +
  geom_line(linewidth=.9) + geom_point(size=2.5) +
  scale_x_continuous(breaks=c(10,20,40,60,80,100), labels=paste0(c(10,20,40,60,80,100),"%")) +
  scale_y_continuous(labels=percent, limits=c(0,1)) +
  labs(title="Overlap of the most strongly connected encoder features",
       x="Training fraction", y="EVN/SVN feature overlap", color=NULL)
```

```{r top-features, eval=has_features}
feature_importance |> filter(training_fraction==100) |>
  group_by(task) |> slice_max(normalized_importance, n=15, with_ties=FALSE) |> ungroup() |>
  mutate(Task=recode(task,evn="EVN",svn="SVN")) |>
  select(Task, `Encoder position`=encoder_position, `Encoder channel`=encoder_channel,
         `Normalized importance`=normalized_importance) |>
  gt(groupname_col="Task") |> tab_header(title="Top 15 latent features in each 100% head") |>
  fmt_number(columns=`Normalized importance`, decimals=6)
```

# Full-training comparison

```{r full}
x |> filter(training_fraction==100) |>
  select(`Trained task`,`Evaluation task`,Mode,Split,
         `Training observations`=n_training_total,
         `Evaluation observations`=n_evaluation_total,AUROC=auroc,AP=auprc) |>
  gt(groupname_col="Trained task") |>
  fmt_integer(columns=c(`Training observations`,`Evaluation observations`)) |>
  fmt_number(columns=c(AUROC,AP),decimals=4)
```

# Reproducibility notes

- Native and cross-task performance use continuous scores, not fixed decisions.
- AlphaGenome uses logits from validation-selected frozen-encoder heads.
- LS-GKM uses continuous gkm-SVM decision scores.
- Every percentage represents an independently trained model.

Rendered at `r format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`.
