---
title: "GRCh38 Native-versus-Cross-Task Evaluation"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
params:
  metrics: "results/cross_task/GRCh38/native_vs_cross_metrics.tsv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=7)
suppressPackageStartupMessages({library(dplyr); library(ggplot2); library(scales); library(gt)})
x <- read.delim(params$metrics, check.names=FALSE) |>
  mutate(Model=recode(model, lsgkm="LS-GKM", alphagenome="AlphaGenome"),
         `Trained task`=recode(model_task, evn="EVN", svn="SVN"),
         `Evaluation task`=recode(evaluation_task, evn="EVN", svn="SVN"),
         Split=factor(recode(split, validation="Validation", test="Test"),
                      levels=c("Validation","Test")),
         Mode=factor(recode(evaluation_mode, native="Native", cross="Cross-task"),
                     levels=c("Native","Cross-task")))
colors <- c(EVN="#0072B2", SVN="#D55E00")
theme_set(theme_minimal(base_size=12) + theme(panel.grid.minor=element_blank(), legend.position="bottom"))
```

# Scope and interpretation

Each learning-curve model is evaluated on its native task and on the other task's
fixed chromosome-matched validation and test cohorts. Cross-task AUROC/AP measures
transfer to the other positive-versus-negative problem; it is not direct
enhancer-versus-silencer classification. EVN and SVN use different synthetic-negative
pools, so native/cross differences can reflect both positive biology and background composition.

# AUROC learning curves

```{r auroc}
ggplot(x, aes(n_training_total, auroc, color=`Evaluation task`, linetype=Mode,
              shape=Mode, group=interaction(`Evaluation task`,Mode))) +
  geom_line(linewidth=.9) + geom_point(size=2.5) +
  facet_grid(Model ~ Split + `Trained task`, scales="free_y") +
  scale_color_manual(values=colors) +
  scale_linetype_manual(values=c(Native="solid", `Cross-task`="dashed")) +
  scale_shape_manual(values=c(Native=16, `Cross-task`=17)) +
  scale_x_continuous(labels=comma) +
  scale_y_continuous(breaks=pretty_breaks(6), expand=expansion(mult=c(.04,.07))) +
  labs(x="Training observations", y="AUROC", color="Evaluation data", linetype="Mode") +
  guides(shape="none",
         linetype=guide_legend(override.aes=list(shape=NA, color="black", linewidth=1)))
```

# Average-precision learning curves

```{r auprc}
ggplot(x, aes(n_training_total, auprc, color=`Evaluation task`, linetype=Mode,
              shape=Mode, group=interaction(`Evaluation task`,Mode))) +
  geom_line(linewidth=.9) + geom_point(size=2.5) +
  facet_grid(Model ~ Split + `Trained task`, scales="free_y") +
  scale_color_manual(values=colors) +
  scale_linetype_manual(values=c(Native="solid", `Cross-task`="dashed")) +
  scale_shape_manual(values=c(Native=16, `Cross-task`=17)) +
  scale_x_continuous(labels=comma) +
  scale_y_continuous(breaks=pretty_breaks(6), expand=expansion(mult=c(.04,.07))) +
  labs(x="Training observations", y="Average Precision", color="Evaluation data", linetype="Mode") +
  guides(shape="none",
         linetype=guide_legend(override.aes=list(shape=NA, color="black", linewidth=1)))
```

# Full-training comparison

```{r full}
x |> filter(training_fraction==100) |>
  select(Model, `Trained task`, `Evaluation task`, Mode, Split,
         `Training observations`=n_training_total,
         `Evaluation observations`=n_evaluation_total, AUROC=auroc, AP=auprc) |>
  gt(groupname_col="Model") |>
  fmt_integer(columns=c(`Training observations`,`Evaluation observations`)) |>
  fmt_number(columns=c(AUROC,AP), decimals=4)
```

# Reproducibility notes

- Native AlphaGenome metrics come from validation-selected checkpoints.
- Cross-task AlphaGenome metrics are computed from exported per-sequence logits.
- LS-GKM metrics use its continuous decision scores.
- All evaluations retain the original task-specific positive and negative labels.

Rendered at `r format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`.
