---
title: "LS-GKM Evaluation Report"
output:
   html_document:
   toc: true
   toc_float: true
   number_sections: true
params:
   results_dir: null
---

## Fold Information
Folds were made using combinations of chromosomes to have a similar loci count between them.

| Fold       | Chromosomes                     | ELS loci          | Missing negs |
| ---------- | ------------------------------- | ----------------- | ------------ |
| 1          | chr1, chr20, chr14, chr22       |             9,744 |           17 |
| 2          | chr2, chr12, chr9, chr13        |             9,583 |           10 |
| 3          | chr17, chr11, chr7, chr15, chrX |             9,995 |           20 |
| 4          | chr6, chr19, chr5, chr4, chr21  |             9,874 |           15 |
| 5          | chr3, chr16, chr10, chr8, chr18 |            10,229 |           15 |

<br/><br/>
An outer represents groups of folds on which a specific fold was left out of training and only used for testing. For instance the "outer1" model was trained using all folds except fold1 and then tested on fold1.

## LS-GKM SVM Params
LS-GKM SVM stands for large-scale, gapped k-mer support vector machine. This model uses a gapped-kmer tool to extract sequence features than can be used to train the vector machine. The following params were used:

 - alphabet size( ${b}$): 4
 - 1X negative synthetic data created from positive regulatory sequences
 - ${\ell}$-mer with truncated function
 - ${\ell}$-mer size (${\ell}$): 11
 - number of informative columns (${k}$): 7
 - max umber of mismatches: 4

The GKM tool creates a matrix M by N, where M is the number of all the possible gapped-kmer combinations, given ${\ell}$ and ${k}$, and N the total number of ungapped possibilites. The number of features depends on ${k}$, ${\ell}$, and ${b}$:

$$M = \binom{\ell}{k} b^k$$

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)

library(dplyr)
library(tidyr)
library(yardstick)
library(ggplot2)
library(gt)

classification_threshold <- 0
```

```{r helper-functions, include=FALSE}
read_scores <- function(file_path) {
  read.delim(
    file_path,
    header = TRUE,
    sep = "\t",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
}


validate_scores <- function(df, file_path) {
  required_columns <- c(
    "chrom",
    "start",
    "end",
    "class",
    "index",
    "score",
    "label",
    "model",
    "task",
    "outer",
    "seq_length"
  )

  missing_columns <- setdiff(
    required_columns,
    colnames(df)
  )

  if (length(missing_columns) > 0) {
    stop(
      "File is missing required columns: ",
      file_path,
      "\nMissing columns: ",
      paste(missing_columns, collapse = ", ")
    )
  }

  if (nrow(df) == 0) {
    stop("File contains no rows: ", file_path)
  }

  if (anyNA(df$score)) {
    stop("Missing score values found in: ", file_path)
  }

  if (anyNA(df$label)) {
    stop("Missing label values found in: ", file_path)
  }

  if (!all(df$label %in% c(0, 1))) {
    stop("Labels must contain only 0 and 1 in: ", file_path)
  }

  if (!all(df$class %in% c("pos", "neg"))) {
    stop(
      "Class must contain only 'pos' and 'neg' in: ",
      file_path
    )
  }

  expected_label <- ifelse(
    df$class == "pos",
    1L,
    0L
  )

  if (!all(df$label == expected_label)) {
    stop(
      "The class and label columns are inconsistent in: ",
      file_path
    )
  }

  if (anyNA(df$start) || anyNA(df$end)) {
    stop("Missing genomic coordinates found in: ", file_path)
  }

  if (any(df$end < df$start)) {
    stop("End coordinate is smaller than start in: ", file_path)
  }

  if (any(df$seq_length != df$end - df$start)) {
    stop("Incorrect sequence lengths found in: ", file_path)
  }

  df
}


prepare_metric_data <- function(df, threshold = 0) {
  df |>
    mutate(
      truth = factor(
        class,
        levels = c("pos", "neg")
      ),
      prediction = factor(
        if_else(score >= threshold, "pos", "neg"),
        levels = c("pos", "neg")
      )
    )
}


calculate_metrics <- function(
  df,
  outer_fold,
  threshold = 0
) {
  metric_data <- prepare_metric_data(
    df,
    threshold = threshold
  )

  tibble(
    outer_fold = outer_fold,
    n_positive = sum(metric_data$label == 1),
    n_negative = sum(metric_data$label == 0),
    positive_prevalence = mean(metric_data$label == 1),

    roc_auc = roc_auc_vec(
      truth = metric_data$truth,
      estimate = metric_data$score,
      event_level = "first"
    ),

    average_precision = average_precision_vec(
      truth = metric_data$truth,
      estimate = metric_data$score,
      event_level = "first"
    ),

    balanced_accuracy = bal_accuracy_vec(
      truth = metric_data$truth,
      estimate = metric_data$prediction,
      event_level = "first"
    ),

    mcc = mcc_vec(
      truth = metric_data$truth,
      estimate = metric_data$prediction
    ),

    sensitivity = sens_vec(
      truth = metric_data$truth,
      estimate = metric_data$prediction,
      event_level = "first"
    ),

    specificity = spec_vec(
      truth = metric_data$truth,
      estimate = metric_data$prediction,
      event_level = "first"
    ),

    precision = precision_vec(
      truth = metric_data$truth,
      estimate = metric_data$prediction,
      event_level = "first"
    ),

    f1 = f_meas_vec(
      truth = metric_data$truth,
      estimate = metric_data$prediction,
      beta = 1,
      event_level = "first"
    )
  )
}
```

```{r read-input, include=FALSE}
results_dir <- params$results_dir

if (is.null(results_dir) || !nzchar(results_dir)) {
  stop(
    "No results directory was provided.\n",
    "Set params$results_dir when rendering the report."
  )
}

if (!dir.exists(results_dir)) {
  stop(
    "Results directory does not exist: ",
    results_dir
  )
}

results_dir <- normalizePath(
  results_dir,
  mustWork = TRUE
)

outer_names <- paste0("outer", 1:5)

outer_files <- file.path(
  results_dir,
  paste0(outer_names, ".tsv")
)

all_outer_file <- file.path(
  results_dir,
  "all_outer.tsv"
)

required_files <- c(
  outer_files,
  all_outer_file
)

missing_files <- required_files[
  !file.exists(required_files)
]

if (length(missing_files) > 0) {
  stop(
    "The following required files are missing:\n",
    paste(missing_files, collapse = "\n")
  )
}

outer_data <- setNames(
  lapply(
    outer_files,
    function(file_path) {
      validate_scores(
        read_scores(file_path),
        file_path
      )
    }
  ),
  outer_names
)

for (outer_name in outer_names) {
  observed_outer <- unique(
    outer_data[[outer_name]]$outer
  )

  if (
    length(observed_outer) != 1 ||
    observed_outer != outer_name
  ) {
    stop(
      "The outer column in ",
      outer_name,
      ".tsv does not match its filename."
    )
  }
}

df_outer_files <- bind_rows(
  outer_data
)

df_all <- validate_scores(
  read_scores(all_outer_file),
  all_outer_file
)

if (nrow(df_outer_files) != nrow(df_all)) {
  stop(
    "The number of rows in all_outer.tsv does not match ",
    "the combined number of rows in outer1.tsv through outer5.tsv."
  )
}

individual_counts <- df_outer_files |>
  count(
    outer,
    class,
    name = "n"
  ) |>
  arrange(
    outer,
    class
  )

consolidated_counts <- df_all |>
  count(
    outer,
    class,
    name = "n"
  ) |>
  arrange(
    outer,
    class
  )

if (!identical(individual_counts, consolidated_counts)) {
  stop(
    "The outer-fold class counts in all_outer.tsv do not match ",
    "the five individual outer-fold files."
  )
}
```

# Dataset composition

The evaluation directory contains five held-out outer-fold prediction files and one consolidated out-of-fold prediction file.

```{r dataset-composition-table}
dataset_composition <- df_all |>
  group_by(outer) |>
  summarise(
    n_total = n(),
    n_positive = sum(label == 1),
    n_negative = sum(label == 0),
    positive_prevalence = mean(label == 1),
    minimum_seq_length = min(seq_length),
    median_seq_length = median(seq_length),
    maximum_seq_length = max(seq_length),
    .groups = "drop"
  ) |>
  arrange(outer)

dataset_composition |>
  gt() |>
  tab_header(
    title = "Dataset composition by outer fold"
  ) |>
  fmt_integer(
    columns = c(
      n_total,
      n_positive,
      n_negative,
      minimum_seq_length,
      median_seq_length,
      maximum_seq_length
    )
  ) |>
  fmt_percent(
    columns = positive_prevalence,
    decimals = 1
  )
```

```{r pooled-dataset-composition}
pooled_composition <- df_all |>
  summarise(
    n_total = n(),
    n_positive = sum(label == 1),
    n_negative = sum(label == 0),
    positive_prevalence = mean(label == 1),
    n_outer_folds = n_distinct(outer),
    minimum_seq_length = min(seq_length),
    median_seq_length = median(seq_length),
    maximum_seq_length = max(seq_length)
  )

pooled_composition |>
  gt() |>
  tab_header(
    title = "Pooled out-of-fold dataset composition"
  ) |>
  fmt_integer(
    columns = c(
      n_total,
      n_positive,
      n_negative,
      n_outer_folds,
      minimum_seq_length,
      median_seq_length,
      maximum_seq_length
    )
  ) |>
  fmt_percent(
    columns = positive_prevalence,
    decimals = 1
  )
```

# Per-fold summary table

Threshold-independent metrics use the continuous LS-GKM score. Threshold-based metrics use a classification threshold of `r classification_threshold`.

```{r calculate-per-fold-metrics}
per_fold_metrics <- bind_rows(
  lapply(
    outer_names,
    function(outer_name) {
      calculate_metrics(
        df = outer_data[[outer_name]],
        outer_fold = outer_name,
        threshold = classification_threshold
      )
    }
  )
)

metric_summary_rows <- per_fold_metrics |>
  summarise(
    across(
      c(
        positive_prevalence,
        roc_auc,
        average_precision,
        balanced_accuracy,
        mcc,
        sensitivity,
        specificity,
        precision,
        f1
      ),
      list(
        mean = mean,
        sd = sd
      ),
      na.rm = TRUE
    )
  )
```

```{r per-fold-summary-table}
per_fold_metrics |>
  gt() |>
  tab_header(
    title = "Performance by outer fold",
    subtitle = paste(
      "Threshold-based metrics use score >=",
      classification_threshold
    )
  ) |>
  fmt_integer(
    columns = c(
      n_positive,
      n_negative
    )
  ) |>
  fmt_percent(
    columns = positive_prevalence,
    decimals = 1
  ) |>
  fmt_number(
    columns = c(
      roc_auc,
      average_precision,
      balanced_accuracy,
      mcc,
      sensitivity,
      specificity,
      precision,
      f1
    ),
    decimals = 3
  )
```

```{r fold-mean-sd-table}
fold_metric_distribution <- per_fold_metrics |>
  summarise(
    across(
      c(
        roc_auc,
        average_precision,
        balanced_accuracy,
        mcc,
        sensitivity,
        specificity,
        precision,
        f1
      ),
      list(
        mean = mean,
        sd = sd
      ),
      na.rm = TRUE
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = c("metric", ".value"),
    names_pattern = "(.+)_(mean|sd)$"
  )

fold_metric_distribution |>
  gt() |>
  tab_header(
    title = "Distribution of performance across outer folds"
  ) |>
  fmt_number(
    columns = c(mean, sd),
    decimals = 3
  )
```

# ROC curves

## ROC curves by outer fold

```{r per-fold-roc-curves}
roc_data <- bind_rows(
  lapply(
    outer_names,
    function(outer_name) {
      metric_data <- prepare_metric_data(
        outer_data[[outer_name]],
        threshold = classification_threshold
      )

      roc_curve(
        metric_data,
        truth = truth,
        score,
        event_level = "first"
      ) |>
        mutate(outer = outer_name)
    }
  )
)

roc_labels <- per_fold_metrics |>
  transmute(
    outer = outer_fold,
    label = paste0(
      "ROC AUC = ",
      sprintf("%.3f", roc_auc)
    )
  )

ggplot(
  roc_data,
  aes(
    x = 1 - specificity,
    y = sensitivity,
    group = outer
  )
) +
  geom_line() +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  geom_text(
    data = roc_labels,
    aes(
      x = 0.30,
      y = 0.08,
      label = label
    ),
    inherit.aes = FALSE,
    hjust = 0
  ) +
  facet_wrap(~outer) +
  coord_equal() +
  labs(
    title = "ROC curves by outer fold",
    x = "False-positive rate",
    y = "True-positive rate"
  )
```

## Pooled out-of-fold ROC curve

```{r pooled-roc-curve}
pooled_metric_data <- prepare_metric_data(
  df_all,
  threshold = classification_threshold
)

pooled_roc_data <- roc_curve(
  pooled_metric_data,
  truth = truth,
  score,
  event_level = "first"
)

pooled_roc_auc <- roc_auc_vec(
  truth = pooled_metric_data$truth,
  estimate = pooled_metric_data$score,
  event_level = "first"
)

ggplot(
  pooled_roc_data,
  aes(
    x = 1 - specificity,
    y = sensitivity
  )
) +
  geom_line() +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  coord_equal() +
  labs(
    title = "Pooled out-of-fold ROC curve",
    subtitle = paste0(
      "ROC AUC = ",
      round(pooled_roc_auc, 3)
    ),
    x = "False-positive rate",
    y = "True-positive rate"
  )
```

# Precision–recall curves

## Precision–recall curves by outer fold

```{r per-fold-pr-curves}
pr_data <- bind_rows(
  lapply(
    outer_names,
    function(outer_name) {
      metric_data <- prepare_metric_data(
        outer_data[[outer_name]],
        threshold = classification_threshold
      )

      pr_curve(
        metric_data,
        truth = truth,
        score,
        event_level = "first"
      ) |>
        mutate(outer = outer_name)
    }
  )
)

fold_prevalence <- df_outer_files |>
  group_by(outer) |>
  summarise(
    prevalence = mean(label == 1),
    .groups = "drop"
  )

pr_labels <- per_fold_metrics |>
  transmute(
    outer = outer_fold,
    label = paste0(
      "AP = ",
      sprintf("%.3f", average_precision),
      "\nPrevalence = ",
      sprintf("%.3f", positive_prevalence)
    )
  )

ggplot(
  pr_data,
  aes(
    x = recall,
    y = precision,
    group = outer
  )
) +
  geom_line() +
  geom_hline(
    data = fold_prevalence,
    aes(yintercept = prevalence),
    linetype = "dashed"
  ) +
  geom_text(
    data = pr_labels,
    aes(
      x = 0.05,
      y = 0.12,
      label = label
    ),
    inherit.aes = FALSE,
    hjust = 0
  ) +
  facet_wrap(~outer) +
  coord_cartesian(
    xlim = c(0, 1),
    ylim = c(0, 1)
  ) +
  labs(
    title = "Precision–recall curves by outer fold",
    subtitle = "Dashed lines indicate positive prevalence",
    x = "Recall",
    y = "Precision"
  )
```

## Pooled out-of-fold precision–recall curve

```{r pooled-pr-curve}
pooled_pr_data <- pr_curve(
  pooled_metric_data,
  truth = truth,
  score,
  event_level = "first"
)

pooled_average_precision <- average_precision_vec(
  truth = pooled_metric_data$truth,
  estimate = pooled_metric_data$score,
  event_level = "first"
)

pooled_prevalence <- mean(
  pooled_metric_data$label == 1
)

ggplot(
  pooled_pr_data,
  aes(
    x = recall,
    y = precision
  )
) +
  geom_line() +
  geom_hline(
    yintercept = pooled_prevalence,
    linetype = "dashed"
  ) +
  coord_cartesian(
    xlim = c(0, 1),
    ylim = c(0, 1)
  ) +
  labs(
    title = "Pooled out-of-fold precision–recall curve",
    subtitle = paste0(
      "Average precision = ",
      round(pooled_average_precision, 3),
      "; positive prevalence = ",
      round(pooled_prevalence, 3)
    ),
    x = "Recall",
    y = "Precision"
  )
```

# Pooled out-of-fold performance

The pooled results combine predictions from all five held-out outer folds. Every observation in this table was scored by a model that did not train on that observation's chromosome group.

```{r pooled-performance}
pooled_metrics <- calculate_metrics(
  df = df_all,
  outer_fold = "all_outer",
  threshold = classification_threshold
)

pooled_metrics |>
  gt() |>
  tab_header(
    title = "Pooled out-of-fold performance",
    subtitle = paste(
      "Threshold-based metrics use score >=",
      classification_threshold
    )
  ) |>
  fmt_integer(
    columns = c(
      n_positive,
      n_negative
    )
  ) |>
  fmt_percent(
    columns = positive_prevalence,
    decimals = 1
  ) |>
  fmt_number(
    columns = c(
      roc_auc,
      average_precision,
      balanced_accuracy,
      mcc,
      sensitivity,
      specificity,
      precision,
      f1
    ),
    decimals = 3
  )
```

```{r pooled-versus-fold-summary}
pooled_fold_comparison <- tibble(
  metric = c(
    "ROC AUC",
    "Average precision",
    "Balanced accuracy",
    "MCC",
    "Sensitivity",
    "Specificity",
    "Precision",
    "F1"
  ),

  pooled = c(
    pooled_metrics$roc_auc,
    pooled_metrics$average_precision,
    pooled_metrics$balanced_accuracy,
    pooled_metrics$mcc,
    pooled_metrics$sensitivity,
    pooled_metrics$specificity,
    pooled_metrics$precision,
    pooled_metrics$f1
  ),

  fold_mean = c(
    mean(per_fold_metrics$roc_auc),
    mean(per_fold_metrics$average_precision),
    mean(per_fold_metrics$balanced_accuracy),
    mean(per_fold_metrics$mcc),
    mean(per_fold_metrics$sensitivity),
    mean(per_fold_metrics$specificity),
    mean(per_fold_metrics$precision),
    mean(per_fold_metrics$f1)
  ),

  fold_sd = c(
    sd(per_fold_metrics$roc_auc),
    sd(per_fold_metrics$average_precision),
    sd(per_fold_metrics$balanced_accuracy),
    sd(per_fold_metrics$mcc),
    sd(per_fold_metrics$sensitivity),
    sd(per_fold_metrics$specificity),
    sd(per_fold_metrics$precision),
    sd(per_fold_metrics$f1)
  )
)

pooled_fold_comparison |>
  gt() |>
  tab_header(
    title = "Pooled performance versus fold-level distribution"
  ) |>
  fmt_number(
    columns = c(
      pooled,
      fold_mean,
      fold_sd
    ),
    decimals = 3
  )
```

# Interpretation

```{r interpretation, results='asis'}
best_roc_fold <- per_fold_metrics |>
  slice_max(
    roc_auc,
    n = 1,
    with_ties = FALSE
  )

lowest_roc_fold <- per_fold_metrics |>
  slice_min(
    roc_auc,
    n = 1,
    with_ties = FALSE
  )

best_ap_fold <- per_fold_metrics |>
  slice_max(
    average_precision,
    n = 1,
    with_ties = FALSE
  )

lowest_ap_fold <- per_fold_metrics |>
  slice_min(
    average_precision,
    n = 1,
    with_ties = FALSE
  )

cat(
  sprintf(
    paste0(
      "The pooled out-of-fold ROC AUC was **%.3f**, while the ",
      "mean ROC AUC across the five outer folds was **%.3f ± %.3f**.\n\n"
    ),
    pooled_metrics$roc_auc,
    mean(per_fold_metrics$roc_auc),
    sd(per_fold_metrics$roc_auc)
  )
)

cat(
  sprintf(
    paste0(
      "The pooled average precision was **%.3f**, compared with a ",
      "positive prevalence of **%.3f**. The mean average precision ",
      "across folds was **%.3f ± %.3f**.\n\n"
    ),
    pooled_metrics$average_precision,
    pooled_metrics$positive_prevalence,
    mean(per_fold_metrics$average_precision),
    sd(per_fold_metrics$average_precision)
  )
)

cat(
  sprintf(
    paste0(
      "ROC AUC ranged from **%.3f** in `%s` to **%.3f** in `%s`. ",
      "Average precision ranged from **%.3f** in `%s` to **%.3f** ",
      "in `%s`.\n\n"
    ),
    lowest_roc_fold$roc_auc,
    lowest_roc_fold$outer_fold,
    best_roc_fold$roc_auc,
    best_roc_fold$outer_fold,
    lowest_ap_fold$average_precision,
    lowest_ap_fold$outer_fold,
    best_ap_fold$average_precision,
    best_ap_fold$outer_fold
  )
)

cat(
  sprintf(
    paste0(
      "At the fixed decision threshold of **%.2f**, pooled sensitivity ",
      "was **%.3f** and pooled specificity was **%.3f**. These values ",
      "describe performance at this specific threshold and should be ",
      "interpreted separately from ROC AUC and average precision.\n\n"
    ),
    classification_threshold,
    pooled_metrics$sensitivity,
    pooled_metrics$specificity
  )
)

cat(
  paste0(
    "Differences between folds may reflect variation in the chromosome ",
    "groups assigned to each held-out fold. Fold-level variability should ",
    "therefore be considered alongside the pooled result rather than relying ",
    "on the pooled result alone.\n"
  )
)
```

# Reproducibility

## Input files

```{r input-files}
tibble(
  file = basename(required_files),
  path = required_files,
  rows = c(
    vapply(
      outer_data,
      nrow,
      integer(1)
    ),
    nrow(df_all)
  )
) |>
  gt() |>
  tab_header(
    title = "Files used to generate this report"
  ) |>
  fmt_integer(
    columns = rows
  )
```

## Session information

```{r session-information}
sessionInfo()
```

