---
title: "LS-GKM GRCh38 Evaluation and Learning-Curve Report"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
params:
  results_dir: null
  training_dir: null
  analysis_name: null
  positive_class: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 8.5, fig.height = 5.5, dpi = 120)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(yardstick)
  library(gt)
  library(scales)
})
classification_threshold <- 0
top_fractions <- c(0.01, 0.05, 0.10)
learning_fractions <- c(10, 20, 40, 60, 80, 100)
performance_y_limits <- c(0.64, 0.73)
ranking_error_y_limits <- c(0.27, 0.33)
outer_names <- paste0("outer", 1:4)
model_colors <- c(outer1 = "#0072B2", outer2 = "#D55E00",
                  outer3 = "#009E73", outer4 = "#CC79A7")
theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom"))
```

```{r helpers, include=FALSE}
read_scores <- function(path) {
  read.delim(path, header = TRUE, sep = "\t", stringsAsFactors = FALSE,
             check.names = FALSE)
}

validate_scores <- function(df, path) {
  required <- c("chrom", "start", "end", "class", "index", "score", "label",
                "split", "model", "task", "dataset", "seq_length")
  missing <- setdiff(required, names(df))
  if (length(missing)) stop("Missing columns in ", path, ": ", paste(missing, collapse = ", "))
  if (!nrow(df)) stop("No rows in ", path)
  if (anyNA(df$score) || anyNA(df$label)) stop("Missing score or label in ", path)
  if (!all(df$label %in% c(0, 1))) stop("Labels must be 0/1 in ", path)
  if (!all(df$class %in% c("pos", "neg"))) stop("Invalid class in ", path)
  if (!all(df$label == ifelse(df$class == "pos", 1L, 0L))) stop("Class/label mismatch in ", path)
  if (!all(df$split %in% c("validation", "test"))) stop("Invalid split in ", path)
  if (anyNA(df$start) || anyNA(df$end) || any(df$end < df$start)) stop("Invalid coordinates in ", path)
  if (any(df$seq_length != df$end - df$start)) stop("Invalid sequence length in ", path)
  df
}

metric_data <- function(df, threshold = classification_threshold) {
  df |>
    mutate(
      truth = factor(if_else(label == 1, "pos", "neg"), levels = c("pos", "neg")),
      prediction = factor(if_else(score >= threshold, "pos", "neg"), levels = c("pos", "neg"))
    )
}

ranking_metrics <- function(df) {
  x <- metric_data(df)
  tibble(
    n = nrow(x),
    n_positive = sum(x$label == 1),
    n_negative = sum(x$label == 0),
    prevalence = mean(x$label == 1),
    auroc = roc_auc_vec(x$truth, x$score, event_level = "first"),
    average_precision = average_precision_vec(x$truth, x$score, event_level = "first")
  )
}

threshold_metrics <- function(df, threshold = classification_threshold) {
  x <- metric_data(df, threshold)
  tibble(
    sensitivity = sens_vec(x$truth, x$prediction, event_level = "first"),
    specificity = spec_vec(x$truth, x$prediction, event_level = "first"),
    precision = precision_vec(x$truth, x$prediction, event_level = "first"),
    f1 = f_meas_vec(x$truth, x$prediction, beta = 1, event_level = "first"),
    balanced_accuracy = bal_accuracy_vec(x$truth, x$prediction, event_level = "first"),
    mcc = mcc_vec(x$truth, x$prediction)
  )
}

count_fasta <- function(path) {
  con <- file(path, open = "r")
  on.exit(close(con))
  total <- 0L
  repeat {
    lines <- readLines(con, n = 100000L, warn = FALSE)
    if (!length(lines)) break
    total <- total + sum(startsWith(lines, ">"))
  }
  total
}

fmt_range <- function(x) sprintf("%.3f–%.3f", min(x), max(x))
```

```{r inputs, include=FALSE}
results_dir <- params$results_dir
training_dir <- params$training_dir
analysis_name <- params$analysis_name
positive_class <- params$positive_class
if (is.null(results_dir) || !nzchar(results_dir)) stop("params$results_dir is required")
if (is.null(training_dir) || !nzchar(training_dir)) stop("params$training_dir is required")
if (is.null(analysis_name) || !nzchar(analysis_name)) stop("params$analysis_name is required")
if (is.null(positive_class) || !nzchar(positive_class)) stop("params$positive_class is required")
results_dir <- normalizePath(results_dir, mustWork = TRUE)
training_dir <- normalizePath(training_dir, mustWork = TRUE)

outer_files <- setNames(file.path(results_dir, paste0(outer_names, ".tsv")), outer_names)
learning_files <- setNames(file.path(results_dir, "learning_curves", paste0(learning_fractions, ".tsv")),
                           as.character(learning_fractions))
required_files <- c(outer_files, learning_files)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files)) stop("Missing report inputs:\n", paste(missing_files, collapse = "\n"))

outer_data <- lapply(outer_files, function(path) validate_scores(read_scores(path), path))
learning_data <- lapply(learning_files, function(path) validate_scores(read_scores(path), path))
for (name in outer_names) {
  if (!identical(unique(outer_data[[name]]$dataset), name)) stop("Dataset mismatch in ", name)
}
for (fraction in names(learning_data)) {
  expected_dataset <- paste0("learning_curve_", fraction)
  if (!identical(unique(learning_data[[fraction]]$dataset), expected_dataset)) {
    stop("Dataset mismatch at ", fraction, "%: expected ", expected_dataset)
  }
}

training_table <- bind_rows(lapply(learning_fractions, function(fraction) {
  pos_file <- file.path(training_dir, paste0("train_", fraction, "_pos.fa"))
  neg_file <- file.path(training_dir, paste0("train_", fraction, "_neg.fa"))
  if (!file.exists(pos_file) || !file.exists(neg_file)) stop("Missing learning-curve FASTA at ", fraction, "%")
  tibble(training_fraction = fraction,
         n_training_positive = count_fasta(pos_file),
         n_training_negative = count_fasta(neg_file)) |>
    mutate(n_training_total = n_training_positive + n_training_negative)
}))

evaluation_reference <- learning_data[["100"]]
input_inventory <- bind_rows(
  tibble(file = unname(outer_files), dataset = names(outer_files), role = "chromosome robustness"),
  tibble(file = unname(learning_files), dataset = paste0(names(learning_files), "%"), role = "learning curve")
) |>
  mutate(number_of_rows = vapply(file, function(path) nrow(read_scores(path)), integer(1)),
         file = sub(paste0("^", results_dir, "/?"), "", file))
```

# Scope and interpretation

The GRCh38 `r analysis_name` experiment contains two complementary designs. Four chromosome-based outer experiments compare models trained with different chromosome compositions. Each has a training set, a validation split used for model-development decisions where applicable, and a held-out test split that remains unseen during training. Their test results are compared descriptively to assess whether chromosome composition produces meaningful performance instability; they are not pooled as an out-of-fold estimate.

The learning-curve experiment is separate from the four outer experiments. It uses one fixed validation set and one fixed test set while varying only the amount of training data through six strictly nested subsets (10%, 20%, 40%, 60%, 80%, and 100%). Thus, these are six models on one fixed learning-curve design, not four-fold learning curves.

```{r dataset-summary}
dataset_summary <- evaluation_reference |>
  summarise(
    `Total observations` = n(),
    `Positive sequences` = sum(label == 1),
    `Negative sequences` = sum(label == 0),
    `Positive prevalence` = mean(label == 1),
    `Minimum sequence length` = min(seq_length),
    `Median sequence length` = median(seq_length),
    `Maximum sequence length` = max(seq_length),
    `Outer sets` = length(outer_names),
    `Learning-curve sizes` = length(learning_fractions)
  )

dataset_summary |>
  gt() |>
  tab_header(title = "Fixed validation and test dataset summary") |>
  fmt_integer(columns = c(`Total observations`, `Positive sequences`, `Negative sequences`,
                          `Minimum sequence length`, `Median sequence length`,
                          `Maximum sequence length`)) |>
  fmt_percent(columns = `Positive prevalence`, decimals = 1)
```

The summary counts the fixed validation and test observations once, using the 100% learning-curve model table; predictions from the other five models are repeated measurements on the same sequences and are not counted as additional observations.

```{r training-size-table}
training_table |>
  rename(`Training fraction (%)` = training_fraction,
         `Training positives` = n_training_positive,
         `Training negatives` = n_training_negative,
         `Total training observations` = n_training_total) |>
  gt() |>
  tab_header(title = "Strictly nested learning-curve training sets") |>
  fmt_integer(columns = c(`Training positives`, `Training negatives`, `Total training observations`))
```

## Model and training configuration

LS-GKM v0.1.2 is a large-scale gapped k-mer support vector machine. It represents DNA sequences through gapped k-mer similarity features and trains an SVM on those sequence-derived features. The number of features depends on ${k}$, ${\ell}$, and ${b}$: 
 
$$M = \binom{\ell}{k} b^k$$

The training command supplied only thread and cache options, so the remaining model settings below are the documented defaults reported by the actual binary.

```{r model-metadata}
model_metadata <- tribble(
  ~Setting, ~Value,
  "LS-GKM binary/version", "../lsgkm/bin/gkmtrain; v0.1.2",
  "Kernel", "Estimated l-mer with truncated filter (-t 2; gkm)",
  "Alphabet size ($b$)", "4 (DNA)",
  "$\\ell$-mer size ($\\ell$)", "11",
  "Number of informative columns ($k$)", "7",
  "Maximum number of mismatches ($d$)", "3",
  "SVM C", "1.0",
  "Convergence epsilon", "0.001",
  "Reverse complements", "Treated as the same feature (-R not set)",
  "Threads", "16 (-T 16)",
  "Kernel cache", "160,000 MB (-m 160000)",
  "Negative:positive ratio", "1:1 target (xfold = 1)",
  "Genome assembly", "GRCh38 / hg38 masked",
  "Negative generation", "gkmSVM::genNullSeqs; nMaxTrials = 500",
  "Learning-curve seed", "42; one shuffle per class, nested prefixes",
  "Negative-generation seed", "Not explicitly set"
)
model_metadata |> gt() |> tab_header(title = "LS-GKM training and dataset-generation configuration")
```

Computational allocation was one node, 16 CPUs, 200 GB RAM, a five-day time limit, and at most five concurrent array tasks. These are execution resources rather than model hyperparameters.

# Learning-curve analysis

Validation and held-out test performance are overlaid below. Absolute training observation counts, rather than percentages alone, define the x-axis. Marginal gains, ranking utility, and final-model summaries remain based on the held-out test split.

```{r learning-metrics, include=FALSE}
learning_split_metrics <- bind_rows(lapply(as.character(learning_fractions), function(fraction) {
  bind_rows(lapply(c("validation", "test"), function(split_name) {
    ranking_metrics(filter(learning_data[[fraction]], split == split_name)) |>
      mutate(training_fraction = as.integer(fraction), split = split_name, .before = 1)
  }))
})) |>
  left_join(training_table, by = "training_fraction") |>
  arrange(split, n_training_total) |>
  mutate(ranking_error = 1 - auroc,
         split = factor(split, levels = c("validation", "test"),
                        labels = c("Validation", "Test")))

learning_metrics <- learning_split_metrics |>
  filter(split == "Test") |>
  arrange(n_training_total)
```

## Predictive performance

```{r learning-auroc}
ggplot(learning_split_metrics,
       aes(n_training_total, auroc, color = split, shape = split, group = split)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  geom_text(data = filter(learning_split_metrics, split == "Test"),
            aes(label = sprintf("%d%%\n%.3f", training_fraction, auroc)),
            vjust = -0.7, size = 3.0, show.legend = FALSE) +
  geom_text(data = filter(learning_split_metrics, split == "Validation"),
            aes(label = sprintf("%d%%\n%.3f", training_fraction, auroc)),
            vjust = 1.7, size = 3.0, show.legend = FALSE) +
  scale_color_manual(values = c(Validation = "#D55E00", Test = "#0072B2")) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.06))) +
  scale_y_continuous(limits = performance_y_limits, breaks = pretty_breaks(n = 6),
                     expand = expansion(mult = c(0.02, 0.02))) +
  labs(title = "Validation and test AUROC by absolute training-set size",
       x = "Training observations", y = "AUROC", color = "Split", shape = "Split")
```

```{r learning-ap}
ggplot(learning_split_metrics,
       aes(n_training_total, average_precision, color = split, shape = split, group = split)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  geom_text(data = filter(learning_split_metrics, split == "Test"),
            aes(label = sprintf("%d%%\n%.3f", training_fraction, average_precision)),
            vjust = -0.7, size = 3.0, show.legend = FALSE) +
  geom_text(data = filter(learning_split_metrics, split == "Validation"),
            aes(label = sprintf("%d%%\n%.3f", training_fraction, average_precision)),
            vjust = 1.7, size = 3.0, show.legend = FALSE) +
  scale_color_manual(values = c(Validation = "#D55E00", Test = "#0072B2")) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.06))) +
  scale_y_continuous(limits = performance_y_limits, breaks = pretty_breaks(n = 6),
                     expand = expansion(mult = c(0.02, 0.02))) +
  labs(title = "Validation and test Average Precision by absolute training-set size",
       x = "Training observations", y = "Average Precision", color = "Split", shape = "Split")
```

## Scaling analysis: AUROC ranking error

AUROC is approximately the probability that a randomly selected positive receives a higher score than a randomly selected negative. Therefore, `1 - AUROC` is the **pairwise AUROC ranking error**: approximately the probability/proportion of positive-negative pairs ordered incorrectly, with ties handled according to the AUROC convention. It is neither mean absolute error nor ordinary thresholded classification error.

```{r ranking-error-linear}
ggplot(learning_split_metrics,
       aes(n_training_total, ranking_error, color = split, shape = split, group = split)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  scale_color_manual(values = c(Validation = "#D55E00", Test = "#0072B2")) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(limits = ranking_error_y_limits, breaks = pretty_breaks(n = 6),
                     expand = expansion(mult = c(0.02, 0.02))) +
  labs(title = "Validation and test pairwise AUROC ranking error",
       x = "Training observations", y = "1 - AUROC", color = "Split", shape = "Split")
```

# Chromosome-composition stability

The outer experiments assess sensitivity to chromosome composition. Validation sets
differ by outer design, whereas every outer model is evaluated on its common held-out
fold 5 test data. These are descriptive stability comparisons, not pooled
cross-validation estimates.

```{r outer-metrics, include=FALSE}
outer_split_metrics <- bind_rows(lapply(outer_names, function(name) {
  bind_rows(lapply(c("validation", "test"), function(split_name) {
    ranking_metrics(filter(outer_data[[name]], split == split_name)) |>
      mutate(outer = name, split = split_name, .before = 1)
  }))
})) |>
  mutate(split = factor(split, levels = c("validation", "test"),
                        labels = c("Validation", "Test")),
         ranking_error = 1 - auroc)

outer_test_metrics <- outer_split_metrics |> filter(split == "Test")
outer_summary <- outer_test_metrics |>
  summarise(mean_auroc = mean(auroc), sd_auroc = sd(auroc),
            min_auroc = min(auroc), max_auroc = max(auroc),
            mean_ap = mean(average_precision), sd_ap = sd(average_precision),
            min_ap = min(average_precision), max_ap = max(average_precision))
```

```{r outer-ranking-plots}
outer_long <- outer_split_metrics |>
  select(outer, split, auroc, average_precision) |>
  pivot_longer(c(auroc, average_precision), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric, auroc = "AUROC",
                         average_precision = "Average Precision"))

ggplot(outer_long,
       aes(outer, value, color = split, shape = split, group = split)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  facet_wrap(~metric) +
  scale_color_manual(values = c(Validation = "#D55E00", Test = "#0072B2")) +
  scale_y_continuous(limits = performance_y_limits, breaks = pretty_breaks(n = 6),
                     expand = expansion(mult = c(0.02, 0.02))) +
  labs(title = "Validation and held-out test ranking performance across outer sets",
       x = "Outer set", y = NULL, color = "Split", shape = "Split")
```

```{r outer-table}
outer_split_metrics |>
  transmute(`Outer set` = outer, Split = split, `Evaluation observations` = n,
            `Positive prevalence` = prevalence, AUROC = auroc,
            `Average Precision` = average_precision,
            `Ranking error` = ranking_error) |>
  gt(groupname_col = "Outer set") |>
  fmt_integer(columns = `Evaluation observations`) |>
  fmt_percent(columns = `Positive prevalence`, decimals = 1) |>
  fmt_number(columns = c(AUROC, `Average Precision`, `Ranking error`), decimals = 4)
```

Across the four outer tests, mean AUROC is
`r sprintf("%.3f ± %.3f", outer_summary$mean_auroc, outer_summary$sd_auroc)`
(range `r fmt_range(outer_test_metrics$auroc)`), and mean AP is
`r sprintf("%.3f ± %.3f", outer_summary$mean_ap, outer_summary$sd_ap)`
(range `r fmt_range(outer_test_metrics$average_precision)`). These descriptive
comparisons do not constitute a statistical significance test.

# Full-training comparison

The table below summarizes threshold-independent validation and test performance for
the model trained on the complete learning-curve training pool. This matches the
primary full-training summary used in the AlphaGenome report.

```{r full-training-metrics}
bind_rows(lapply(c("validation", "test"), function(split_name) {
  ranking_metrics(filter(learning_data[["100"]], split == split_name)) |>
    mutate(Split = if_else(split_name == "validation", "Validation", "Test"),
           .before = 1)
})) |>
  transmute(Split, `Evaluation observations` = n,
            `Positive prevalence` = prevalence,
            AUROC = auroc, `Average Precision` = average_precision,
            `Ranking error` = 1 - auroc) |>
  gt() |>
  tab_header(title = "100% LS-GKM model performance") |>
  fmt_integer(columns = `Evaluation observations`) |>
  fmt_percent(columns = `Positive prevalence`, decimals = 1) |>
  fmt_number(columns = c(AUROC, `Average Precision`, `Ranking error`), decimals = 4)
```

# Reproducibility

## Input files

```{r input-inventory}
input_inventory |>
  rename(File = file, Dataset = dataset, Role = role, `Number of rows` = number_of_rows) |>
  gt() |>
  fmt_integer(columns = `Number of rows`)
```

## Analysis configuration

```{r analysis-config}
tribble(
  ~Setting, ~Value,
  "Results directory", results_dir,
  "Training-data directory", training_dir,
  "Threshold", as.character(classification_threshold),
  "Learning-curve fractions", paste0(paste(learning_fractions, collapse = ", "), "%"),
  "Positive class", paste0(positive_class, " (pos; label 1)")
) |> gt()
```

Rendered at `r format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`.

## Software environment

```{r session-info}
sessionInfo()
```

