---
title: "LS-GKM GRCh38 Enhancer-versus-Silencer Comparison"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
params:
  evn_results_dir: null
  svn_results_dir: null
  evn_training_dir: null
  svn_training_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5.5, dpi = 120)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(yardstick)
  library(gt)
  library(scales)
})

learning_fractions <- c(10, 20, 40, 60, 80, 100)
performance_y_limits <- c(0.64, 0.73)
ranking_error_y_limits <- c(0.27, 0.33)
task_colors <- c("Enhancer vs negative" = "#0072B2",
                 "Silencer vs negative" = "#D55E00")

theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom"))
```

```{r helpers, include=FALSE}
read_scores <- function(path, expected_dataset) {
  required <- c("chrom", "start", "end", "class", "index", "score", "label",
                "split", "model", "task", "dataset", "seq_length")
  df <- read.delim(path, header = TRUE, sep = "\t", stringsAsFactors = FALSE,
                   check.names = FALSE)
  missing <- setdiff(required, names(df))
  if (length(missing)) stop("Missing columns in ", path, ": ", paste(missing, collapse = ", "))
  if (!nrow(df)) stop("No rows in ", path)
  if (anyNA(df$score) || anyNA(df$label)) stop("Missing score or label in ", path)
  if (!all(df$label %in% c(0, 1))) stop("Labels must be 0/1 in ", path)
  if (!all(df$class %in% c("pos", "neg"))) stop("Invalid class in ", path)
  if (!all(df$label == ifelse(df$class == "pos", 1L, 0L))) stop("Class/label mismatch in ", path)
  if (!setequal(unique(df$split), c("validation", "test"))) {
    stop("Expected validation and test rows in ", path)
  }
  if (!identical(unique(df$model), "lsgkm")) stop("Model mismatch in ", path)
  if (!identical(unique(df$task), "GRCh38")) stop("Task mismatch in ", path)
  if (!identical(unique(df$dataset), expected_dataset)) {
    stop("Dataset mismatch in ", path, ": expected ", expected_dataset)
  }
  if (anyNA(df$start) || anyNA(df$end) || any(df$end < df$start)) {
    stop("Invalid coordinates in ", path)
  }
  if (any(df$seq_length != df$end - df$start)) stop("Invalid sequence length in ", path)
  df
}

count_fasta <- function(path) {
  if (!file.exists(path)) stop("Missing training FASTA: ", path)
  con <- file(path, open = "r")
  on.exit(close(con))
  total <- 0L
  repeat {
    lines <- readLines(con, n = 100000L, warn = FALSE)
    if (!length(lines)) break
    total <- total + sum(startsWith(lines, ">"))
  }
  total
}

ranking_metrics <- function(df) {
  truth <- factor(ifelse(df$label == 1, "pos", "neg"), levels = c("pos", "neg"))
  tibble(
    n = nrow(df),
    prevalence = mean(df$label == 1),
    auroc = roc_auc_vec(truth, df$score, event_level = "first"),
    average_precision = average_precision_vec(truth, df$score, event_level = "first")
  )
}
```

```{r inputs, include=FALSE}
task_specs <- tibble(
  comparison = c("Enhancer vs negative", "Silencer vs negative"),
  results_dir = c(params$evn_results_dir, params$svn_results_dir),
  training_dir = c(params$evn_training_dir, params$svn_training_dir)
)

if (any(vapply(task_specs$results_dir, is.null, logical(1))) ||
    any(vapply(task_specs$training_dir, is.null, logical(1)))) {
  stop("All four report directory parameters are required")
}
task_specs$results_dir <- vapply(task_specs$results_dir, normalizePath,
                                 character(1), mustWork = TRUE)
task_specs$training_dir <- vapply(task_specs$training_dir, normalizePath,
                                  character(1), mustWork = TRUE)

score_data <- bind_rows(lapply(seq_len(nrow(task_specs)), function(i) {
  bind_rows(lapply(learning_fractions, function(fraction) {
    path <- file.path(task_specs$results_dir[[i]], "learning_curves",
                      paste0(fraction, ".tsv"))
    if (!file.exists(path)) stop("Missing score table: ", path)
    read_scores(path, paste0("learning_curve_", fraction)) |>
      mutate(comparison = task_specs$comparison[[i]],
             training_fraction = fraction,
             source_file = path)
  }))
}))

training_sizes <- bind_rows(lapply(seq_len(nrow(task_specs)), function(i) {
  bind_rows(lapply(learning_fractions, function(fraction) {
    pos_file <- file.path(task_specs$training_dir[[i]], paste0("train_", fraction, "_pos.fa"))
    neg_file <- file.path(task_specs$training_dir[[i]], paste0("train_", fraction, "_neg.fa"))
    n_positive <- count_fasta(pos_file)
    n_negative <- count_fasta(neg_file)
    tibble(comparison = task_specs$comparison[[i]], training_fraction = fraction,
           n_training_positive = n_positive, n_training_negative = n_negative,
           n_training_total = n_positive + n_negative)
  }))
}))

split_metrics <- score_data |>
  group_by(comparison, training_fraction, split) |>
  group_modify(~ ranking_metrics(.x)) |>
  ungroup() |>
  left_join(training_sizes, by = c("comparison", "training_fraction")) |>
  mutate(ranking_error = 1 - auroc,
         comparison = factor(comparison, levels = names(task_colors)),
         split = factor(split, levels = c("validation", "test"),
                        labels = c("Validation", "Test"))) |>
  arrange(split, comparison, n_training_total)

outer_score_data <- bind_rows(lapply(seq_len(nrow(task_specs)), function(i) {
  bind_rows(lapply(paste0("outer", 1:4), function(outer_name) {
    path <- file.path(task_specs$results_dir[[i]], paste0(outer_name, ".tsv"))
    if (!file.exists(path)) stop("Missing outer score table: ", path)
    read_scores(path, outer_name) |>
      mutate(comparison = task_specs$comparison[[i]], outer = outer_name,
             source_file = path)
  }))
}))

outer_split_metrics <- outer_score_data |>
  group_by(comparison, outer, split) |>
  group_modify(~ ranking_metrics(.x)) |>
  ungroup() |>
  mutate(ranking_error = 1 - auroc,
         comparison = factor(comparison, levels = names(task_colors)),
         outer = factor(outer, levels = paste0("outer", 1:4)),
         split = factor(split, levels = c("validation", "test"),
                        labels = c("Validation", "Test")))
```

# Scope and interpretation

This report compares how LS-GKM performance scales with training-set size for two
binary tasks: enhancer versus matched synthetic negative sequences (EVN), and
silencer versus matched synthetic negative sequences (SVN). The same metric
definitions are used in both analyses.

The comparison is descriptive rather than a paired benchmark. EVN and SVN use
different positive classes and different validation/test loci, so a vertical
difference between their curves may reflect both biological task difficulty and
evaluation-cohort composition. Absolute training-observation counts are therefore
used on the x-axis instead of treating equal percentages as equal sample sizes.

```{r training-sizes}
training_sizes |>
  mutate(`Training fraction` = paste0(training_fraction, "%")) |>
  select(Task = comparison, `Training fraction`,
         `Training positives` = n_training_positive,
         `Training negatives` = n_training_negative,
         `Total training observations` = n_training_total) |>
  gt(groupname_col = "Task") |>
  tab_header(title = "Learning-curve training-set sizes") |>
  fmt_integer(columns = c(`Training positives`, `Training negatives`,
                          `Total training observations`))
```

# Learning curves

## AUROC

AUROC summarizes score ranking across all possible thresholds. Validation and test
results are shown in separate panels so that generalization trends can be compared
without conflating the two evaluation splits.

```{r auroc-curves}
ggplot(split_metrics,
       aes(n_training_total, auroc, color = comparison, shape = comparison,
           group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~split, ncol = 2) +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(limits = performance_y_limits, breaks = pretty_breaks(n = 6),
                     expand = expansion(mult = c(0.02, 0.02))) +
  labs(title = "AUROC learning curves",
       x = "Training observations", y = "AUROC", color = NULL, shape = NULL)
```

## Average precision

Average precision summarizes positive-class ranking and complements AUROC.

```{r ap-curves}
ggplot(split_metrics,
       aes(n_training_total, average_precision, color = comparison,
           shape = comparison, group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~split, ncol = 2) +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(limits = performance_y_limits, breaks = pretty_breaks(n = 6),
                     expand = expansion(mult = c(0.02, 0.02))) +
  labs(title = "Average-precision learning curves",
       x = "Training observations", y = "Average Precision",
       color = NULL, shape = NULL)
```

## Pairwise AUROC ranking error

`1 - AUROC` is the pairwise AUROC ranking error: approximately the probability that
a positive-negative pair is ordered incorrectly, with ties handled by the AUROC
convention. It is not thresholded classification error.

```{r ranking-error-curves}
ggplot(split_metrics,
       aes(n_training_total, ranking_error, color = comparison,
           shape = comparison, group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~split, ncol = 2) +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(limits = ranking_error_y_limits, breaks = pretty_breaks(n = 6),
                     expand = expansion(mult = c(0.02, 0.02))) +
  labs(title = "Pairwise AUROC ranking-error curves",
       x = "Training observations", y = "1 - AUROC",
       color = NULL, shape = NULL)
```

# Chromosome-composition stability

Outer-set numbering is task-specific because chromosome groupings were balanced
separately for EVN and SVN. These plots compare task-specific stability patterns,
not identical chromosome compositions.

```{r outer-stability}
outer_long <- outer_split_metrics |>
  select(comparison, outer, split, auroc, average_precision) |>
  pivot_longer(c(auroc, average_precision), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric, auroc = "AUROC",
                         average_precision = "Average Precision"))

ggplot(outer_long,
       aes(outer, value, color = comparison, shape = comparison,
           group = comparison)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.8) +
  facet_grid(split ~ metric) +
  scale_color_manual(values = task_colors) +
  scale_y_continuous(limits = performance_y_limits, breaks = pretty_breaks(n = 6),
                     expand = expansion(mult = c(0.02, 0.02))) +
  labs(title = "Validation and held-out test stability",
       x = "Task-specific outer set", y = NULL, color = NULL, shape = NULL)
```

```{r outer-summary}
outer_split_metrics |>
  filter(split == "Test") |>
  group_by(comparison) |>
  summarise(`Mean AUROC` = mean(auroc), `SD AUROC` = sd(auroc),
            `Minimum AUROC` = min(auroc), `Maximum AUROC` = max(auroc),
            `Mean AP` = mean(average_precision), `SD AP` = sd(average_precision),
            `Minimum AP` = min(average_precision),
            `Maximum AP` = max(average_precision), .groups = "drop") |>
  rename(Task = comparison) |>
  gt() |>
  tab_header(title = "Held-out outer-test summary") |>
  fmt_number(columns = where(is.numeric), decimals = 4)
```

# Full-training comparison

```{r final-summary}
split_metrics |>
  filter(training_fraction == 100) |>
  transmute(Task = comparison, Split = split,
            `Training observations` = n_training_total,
            `Evaluation observations` = n,
            `Positive prevalence` = prevalence,
            AUROC = auroc, `Average Precision` = average_precision,
            `Ranking error` = ranking_error) |>
  gt(groupname_col = "Task") |>
  tab_header(title = "100% training-set performance") |>
  fmt_integer(columns = c(`Training observations`, `Evaluation observations`)) |>
  fmt_percent(columns = `Positive prevalence`, decimals = 1) |>
  fmt_number(columns = c(AUROC, `Average Precision`, `Ranking error`), decimals = 4)
```

# Reproducibility notes

- The report reads the six merged learning-curve and four merged outer TSV files for each task.
- The `score` column is the continuous LS-GKM decision score; no threshold is used
  for AUROC, average precision, or ranking error.
- Training sizes are counted directly from each positive and negative FASTA.
- Outer-set numbering is task-specific; stability comparisons are descriptive.

