---
title: "LS-GKM GRCh38 Enhancer-versus-Silencer Comparison"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
params:
  evn_results_dir: null
  svn_results_dir: null
  evn_training_dir: null
  svn_training_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5.5, dpi = 120)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(yardstick)
  library(gt)
  library(scales)
})

learning_fractions <- c(10, 20, 40, 60, 80, 100)
top_fractions <- c(0.01, 0.05, 0.10)
task_colors <- c("Enhancer vs negative" = "#0072B2",
                 "Silencer vs negative" = "#D55E00")

theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom"))
```

```{r helpers, include=FALSE}
read_scores <- function(path, expected_dataset) {
  required <- c("chrom", "start", "end", "class", "index", "score", "label",
                "split", "model", "task", "dataset", "seq_length")
  df <- read.delim(path, header = TRUE, sep = "\t", stringsAsFactors = FALSE,
                   check.names = FALSE)
  missing <- setdiff(required, names(df))
  if (length(missing)) stop("Missing columns in ", path, ": ", paste(missing, collapse = ", "))
  if (!nrow(df)) stop("No rows in ", path)
  if (anyNA(df$score) || anyNA(df$label)) stop("Missing score or label in ", path)
  if (!all(df$label %in% c(0, 1))) stop("Labels must be 0/1 in ", path)
  if (!all(df$class %in% c("pos", "neg"))) stop("Invalid class in ", path)
  if (!all(df$label == ifelse(df$class == "pos", 1L, 0L))) stop("Class/label mismatch in ", path)
  if (!setequal(unique(df$split), c("validation", "test"))) {
    stop("Expected validation and test rows in ", path)
  }
  if (!identical(unique(df$model), "lsgkm")) stop("Model mismatch in ", path)
  if (!identical(unique(df$task), "GRCh38")) stop("Task mismatch in ", path)
  if (!identical(unique(df$dataset), expected_dataset)) {
    stop("Dataset mismatch in ", path, ": expected ", expected_dataset)
  }
  if (anyNA(df$start) || anyNA(df$end) || any(df$end < df$start)) {
    stop("Invalid coordinates in ", path)
  }
  if (any(df$seq_length != df$end - df$start)) stop("Invalid sequence length in ", path)
  df
}

count_fasta <- function(path) {
  if (!file.exists(path)) stop("Missing training FASTA: ", path)
  con <- file(path, open = "r")
  on.exit(close(con))
  total <- 0L
  repeat {
    lines <- readLines(con, n = 100000L, warn = FALSE)
    if (!length(lines)) break
    total <- total + sum(startsWith(lines, ">"))
  }
  total
}

ranking_metrics <- function(df) {
  truth <- factor(ifelse(df$label == 1, "pos", "neg"), levels = c("pos", "neg"))
  tibble(
    n = nrow(df),
    prevalence = mean(df$label == 1),
    auroc = roc_auc_vec(truth, df$score, event_level = "first"),
    average_precision = average_precision_vec(truth, df$score, event_level = "first")
  )
}
```

```{r inputs, include=FALSE}
task_specs <- tibble(
  comparison = c("Enhancer vs negative", "Silencer vs negative"),
  results_dir = c(params$evn_results_dir, params$svn_results_dir),
  training_dir = c(params$evn_training_dir, params$svn_training_dir)
)

if (any(vapply(task_specs$results_dir, is.null, logical(1))) ||
    any(vapply(task_specs$training_dir, is.null, logical(1)))) {
  stop("All four report directory parameters are required")
}
task_specs$results_dir <- vapply(task_specs$results_dir, normalizePath,
                                 character(1), mustWork = TRUE)
task_specs$training_dir <- vapply(task_specs$training_dir, normalizePath,
                                  character(1), mustWork = TRUE)

score_data <- bind_rows(lapply(seq_len(nrow(task_specs)), function(i) {
  bind_rows(lapply(learning_fractions, function(fraction) {
    path <- file.path(task_specs$results_dir[[i]], "learning_curves",
                      paste0(fraction, ".tsv"))
    if (!file.exists(path)) stop("Missing score table: ", path)
    read_scores(path, paste0("learning_curve_", fraction)) |>
      mutate(comparison = task_specs$comparison[[i]],
             training_fraction = fraction,
             source_file = path)
  }))
}))

training_sizes <- bind_rows(lapply(seq_len(nrow(task_specs)), function(i) {
  bind_rows(lapply(learning_fractions, function(fraction) {
    pos_file <- file.path(task_specs$training_dir[[i]], paste0("train_", fraction, "_pos.fa"))
    neg_file <- file.path(task_specs$training_dir[[i]], paste0("train_", fraction, "_neg.fa"))
    n_positive <- count_fasta(pos_file)
    n_negative <- count_fasta(neg_file)
    tibble(comparison = task_specs$comparison[[i]], training_fraction = fraction,
           n_training_positive = n_positive, n_training_negative = n_negative,
           n_training_total = n_positive + n_negative)
  }))
}))

split_metrics <- score_data |>
  group_by(comparison, training_fraction, split) |>
  group_modify(~ ranking_metrics(.x)) |>
  ungroup() |>
  left_join(training_sizes, by = c("comparison", "training_fraction")) |>
  mutate(ranking_error = 1 - auroc,
         comparison = factor(comparison, levels = names(task_colors)),
         split = factor(split, levels = c("validation", "test"),
                        labels = c("Validation", "Test"))) |>
  arrange(split, comparison, n_training_total)

top_ranking <- score_data |>
  filter(split == "test") |>
  group_by(comparison, training_fraction) |>
  group_modify(function(.x, .y) {
    ranked <- arrange(.x, desc(score))
    prevalence <- mean(ranked$label == 1)
    bind_rows(lapply(top_fractions, function(top_fraction) {
      n_selected <- min(nrow(ranked), max(1L, ceiling(nrow(ranked) * top_fraction)))
      selected <- slice_head(ranked, n = n_selected)
      precision <- mean(selected$label == 1)
      tibble(top_fraction = top_fraction, n_selected = n_selected,
             precision = precision, prevalence = prevalence,
             enrichment = precision / prevalence)
    }))
  }) |>
  ungroup() |>
  left_join(training_sizes, by = c("comparison", "training_fraction")) |>
  mutate(comparison = factor(comparison, levels = names(task_colors)),
         rank_cutoff = factor(top_fraction, levels = top_fractions,
                              labels = paste("Top", percent(top_fractions, accuracy = 1))))
```

# Scope and interpretation

This report compares how LS-GKM performance scales with training-set size for two
binary tasks: enhancer versus matched synthetic negative sequences (EVN), and
silencer versus matched synthetic negative sequences (SVN). The same metric
definitions are used in both analyses.

The comparison is descriptive rather than a paired benchmark. EVN and SVN use
different positive classes and different validation/test loci, so a vertical
difference between their curves may reflect both biological task difficulty and
evaluation-cohort composition. Absolute training-observation counts are therefore
used on the x-axis instead of treating equal percentages as equal sample sizes.

```{r training-sizes}
training_sizes |>
  mutate(`Training fraction` = paste0(training_fraction, "%")) |>
  select(Task = comparison, `Training fraction`,
         `Training positives` = n_training_positive,
         `Training negatives` = n_training_negative,
         `Total training observations` = n_training_total) |>
  gt(groupname_col = "Task") |>
  tab_header(title = "Learning-curve training-set sizes") |>
  fmt_integer(columns = c(`Training positives`, `Training negatives`,
                          `Total training observations`))
```

# AUROC learning curves

AUROC summarizes score ranking across all possible thresholds. Validation and test
results are shown in separate panels so that generalization trends can be compared
without conflating the two evaluation splits.

```{r auroc-curves}
ggplot(split_metrics,
       aes(n_training_total, auroc, color = comparison, shape = comparison,
           group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~split, ncol = 2, scales = "free_y") +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.14, 0.18))) +
  labs(title = "AUROC learning curves",
       x = "Training observations", y = "AUROC", color = NULL, shape = NULL)
```

# Average-precision learning curves

Average precision summarizes positive-class ranking and complements AUROC. It is
shown because enrichment among the highest-ranked loci is a central downstream use
of these models.

```{r ap-curves}
ggplot(split_metrics,
       aes(n_training_total, average_precision, color = comparison,
           shape = comparison, group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~split, ncol = 2, scales = "free_y") +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.14, 0.18))) +
  labs(title = "Average-precision learning curves",
       x = "Training observations", y = "Average Precision",
       color = NULL, shape = NULL)
```

# Pairwise AUROC ranking error

`1 - AUROC` is the pairwise AUROC ranking error: approximately the probability that
a positive-negative pair is ordered incorrectly, with ties handled by the AUROC
convention. It is not thresholded classification error.

```{r ranking-error-curves}
ggplot(split_metrics,
       aes(n_training_total, ranking_error, color = comparison,
           shape = comparison, group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~split, ncol = 2, scales = "free_y") +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.14, 0.18))) +
  labs(title = "Pairwise AUROC ranking-error curves",
       x = "Training observations", y = "1 - AUROC",
       color = NULL, shape = NULL)
```

# Highest-ranked test-locus enrichment

For each model, test loci are ordered by decreasing LS-GKM score. Enrichment is the
positive-class precision within the selected top fraction divided by the positive
prevalence of that task's complete test set.

```{r top-enrichment-curves}
ggplot(top_ranking,
       aes(n_training_total, enrichment, color = comparison,
           shape = comparison, group = comparison)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~rank_cutoff, nrow = 1, scales = "free_y") +
  scale_color_manual(values = task_colors) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0.03, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.10, 0.14))) +
  labs(title = "Enrichment of positives among highest-ranked test loci",
       x = "Training observations", y = "Enrichment over test prevalence",
       color = NULL, shape = NULL)
```

# Full-training comparison

```{r final-summary}
split_metrics |>
  filter(training_fraction == 100) |>
  transmute(Task = comparison, Split = split,
            `Training observations` = n_training_total,
            `Evaluation observations` = n,
            `Positive prevalence` = prevalence,
            AUROC = auroc, `Average Precision` = average_precision,
            `Ranking error` = ranking_error) |>
  gt(groupname_col = "Task") |>
  tab_header(title = "100% training-set performance") |>
  fmt_integer(columns = c(`Training observations`, `Evaluation observations`)) |>
  fmt_percent(columns = `Positive prevalence`, decimals = 1) |>
  fmt_number(columns = c(AUROC, `Average Precision`, `Ranking error`), decimals = 4)
```

# Reproducibility notes

- The report reads the six merged learning-curve TSV files for each task.
- The `score` column is the continuous LS-GKM decision score; no threshold is used
  for AUROC, average precision, ranking error, or enrichment.
- Training sizes are counted directly from each positive and negative FASTA.
- Top-fraction sizes use `ceiling(N × fraction)`, are capped at `N`, and contain at
  least one sequence, matching the task-specific LS-GKM reports.

